---
title: "DESeq2_Mus_musculus"
author: "Aleksandra Śmigas"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Report on performing differential expression analysis using the DESeq2 package.

## Introduction
Packages used in the analysis:

* dplyr
* tidyr
* vroom
* DESeq2
* apeglm
* ggplot2
* pheatmap

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(vroom)
library(DESeq2)
library(apeglm)
library(ggplot2)
library(pheatmap)
library(DT)
library(rmarkdown)
```

**metadata** table preview
```{r, echo=FALSE, message=FALSE}
metadata <- vroom("~/Desktop/RNAseq_beginning/data/SraRunTable.txt")
paged_table(metadata)
```

**row_counts** table preview - head
```{r, echo=FALSE, message=FALSE}
counts_file <- "~/Desktop/RNAseq_beginning/data/counts_all.count"

raw_counts <-
  vroom(counts_file, col_names = TRUE) %>% tibble::column_to_rownames(., var = "gene_name")
datatable(head(raw_counts), class = 'cell-border stripe')

```

**coldata** table 
```{r, echo=FALSE, message=FALSE}
coldata <- cbind(
  c("chronic_intermittent_hypoxia","chronic_intermittent_hypoxia","chronic_intermittent_hypoxia",
    "control","control","control")
)

rownames(coldata) <- colnames(raw_counts)
colnames(coldata) <- c("treatment")
coldata <- data.frame(coldata)
coldata$treatment <- as.factor(coldata$treatment)
datatable(coldata,class = 'cell-border stripe') 
```

## Pre-filtering
Removal of genes for which the sum of counts in all samples was less than 10
```{r, echo=FALSE}
genes_raw <- nrow(raw_counts)
```

```{r}
raw_counts <- raw_counts %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  filter(total >= 10) %>%
  select(!total)
```

```{r, echo=FALSE}
genes_raw_filtered <- nrow(raw_counts)
```

Number of genes before filtration: **`r genes_raw`**.

Number of genes after filtration: **`r genes_raw_filtered`**.

## Unsupervised clustering analysis

```{r, message=FALSE}
dds <- DESeqDataSetFromMatrix(countData = raw_counts,
                              colData = coldata,
                              design = ~treatment)
dds <- DESeq(dds)

# qc
vsd <- vst(dds, blind=TRUE) #log transformation of counts
vsd_mat <- assay(vsd)
vsd_cor <- cor(vsd_mat)
pheatmap(vsd_cor, annotation = select(coldata,treatment))

plotPCA(vsd, intgroup="treatment")

```

## Differential expression analysis
```{r, message=FALSE}
res <- results(dds) 
resLFC <- lfcShrink(dds, coef="treatment_control_vs_chronic_intermittent_hypoxia", type="apeglm")

plotMA(res, ylim=c(-2,2))
plotMA(resLFC, ylim=c(-2,2))
```
