---
title: "edgeR_Mus_musculus"
author: "Aleksandra Śmigas"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Report on performing differential expression analysis using the edgeR package.

# Introduction
## Packages
Packages used in the analysis:

* edgeR
* ggplot2
* vroom
* dplyr
* tidyr
* DT
* rmarkdown
* statmod

```{r, echo=FALSE, message=FALSE}
library(edgeR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(vroom)
library(DT)
library(rmarkdown)
library(statmod)
```
## Metadata
**metadata** table preview
```{r, echo=FALSE, message=FALSE}
metadata <- vroom("~/Desktop/RNAseq_beginning/data/SraRunTable.txt")
paged_table(metadata)
```
## Raw counts
**row_counts** table preview - head. Deletion of control samples (SRR20753642,SRR20753639) - incorrect grouping of control data.
```{r, echo=FALSE, message=FALSE}
counts_file <- "~/Desktop/RNAseq_beginning/data/counts_all.count"

raw_counts <- vroom(counts_file, col_names = TRUE) %>% tibble::column_to_rownames(., var = "gene_name")
raw_counts <- raw_counts %>% select(!c(SRR20753642,SRR20753639))
datatable(head(raw_counts),class = 'cell-border stripe')
```

# Analysis
## Creating DGEList object.
**DGEList object** preview - sample 
```{r, echo=FALSE, message=FALSE}

group <- as.factor(c("cih", "cih", "control", "control"))

y <- DGEList(counts=raw_counts,group=group)
datatable(y$samples,class = 'cell-border stripe')

```

## Filterning
Smallest group
```{r,echp=FALSE}
table(y$samples$group)
```

**minimum library size** - 0.6 (mistake)

```{r}
L = 0.6 # the minimum library size in millions.
min_cpm_0.6 <- 10/L
min_cpm_0.6
```

```{r, echo=FALSE}
min_s = min(table(y$samples$group)) # the smallest group sample size 

keep_0.6 <- rowSums(cpm(y) > min_cpm_0.6) >= min_s 
datatable(as.data.frame(table(keep_0.6)),class = 'cell-border stripe')
```

**minimum library size** - 66

```{r}
L = 66 # the minimum library size in millions.
min_cpm_66 <- 10/L
min_cpm_66
```

```{r, echo=FALSE}
min_s <- min(table(y$samples$group)) # the smallest group sample size 

keep_66 <- rowSums(cpm(y) > min_cpm_66) >= min_s 
datatable(as.data.frame(table(keep_66)),class = 'cell-border stripe')
```

Samples table
```{r, echo=FALSE}
y <- y[keep_66, , keep.lib.sizes=FALSE]
```

## Normalization
Samples table after normalization
```{r}
y <- calcNormFactors(y)
```

```{r, echo=FALSE}
datatable(y$samples,class = 'cell-border stripe')
```

## Design matrix
```{r}
design <- model.matrix(~0+group)
colnames(design) <- levels(group)
```

```{r, echo=FALSE}
datatable(as.data.frame(design),class = 'cell-border stripe')
```

## Estimation of dispersion
```{r, echo=FALSE}
y <- estimateDisp(y, design, robust=TRUE)
plotBCV(y)
```

```{r, echo=FALSE}
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
```



# Differential expression analysis
```{r, echo=FALSE}
cih.v.control <- makeContrasts(cih-control, levels=design)
res <- glmQLFTest(fit, contrast=cih.v.control)
```

Contrast table
```{r, echo=FALSE}
datatable(cih.v.control,class = 'cell-border stripe')
```

Head of top_DE
```{r, echo=FALSE}
#top DE genes
top_DE <- topTags(res, n=Inf)
datatable(head(top_DE$table),class = 'cell-border stripe')
```

```{r, echo=FALSE}
is.de <- decideTestsDGE(res)

datatable(summary(is.de),class = 'cell-border stripe')


plotMD(res, status=is.de, values=c(1,-1), col=c("red","blue"),legend="topright")

plot_data <- data.frame(AveLogCPM=res$AveLogCPM, logFC=res$table$logFC, status = as.factor(is.de))
ggplot(plot_data) + geom_point(aes(x=AveLogCPM,y=logFC,color = status)) + xlab("Average log CPM") + ylab("log-fold-change") +  theme(axis.text=element_text(size=12), axis.title=element_text(size=14,face="bold")) + ggtitle(colnames(is.de)) + scale_color_manual(values=c("blue", "black", "red")) + theme_classic()  

plot(x=top_DE$table$logFC, y=-log(top_DE$table$FDR), pch=20, xlab="logFC", ylab="-log FDR")
legend("bottomleft", legend=c("Up-regulated", "Down-regulated"),col=c("steelblue","aquamarine3"),cex=0.9,pch=20,box.lty=0)
title("cih vs control")
abline(h=-log(0.05), col="red", lty=2)
with(subset(top_DE$table, FDR<.05 & logFC>1), points(logFC, -log(FDR), pch=20, col="steelblue",lty=2))
with(subset(top_DE$table, FDR<.05 & logFC<(-1)), points(logFC, -log(FDR), pch=20, col="aquamarine3",lty=2))
abline(v=1, col="red", lty=2)
abline(v=-1, col="red", lty=2)

```


